---
title: "OCN683_homework_1"
author: "kelli konicek"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Dr. Edwards, I hope this is my worst homework. Or, at least, I hope this *isn't* my best homework. 

# Load Libraries
```{r}
library(tidyverse)
library(here)
library(ggeffects)
library(GGally)
library(effects)
library(ggResidpanel)
library(car)
```

# Upload Data, remove species not mentioned in prompt(?)
I am a little concerned I am making a bad mistake, but the 'oma'o and the elepaio that weren't mentioned in the prompt and also do not have very many observations, so I am going to assume they are left out on purpose.
```{r}
raw <- read_csv(here("week_1_forest_birds", "data","foraging.height.edit.csv")) %>%
  filter(SPECIES != "OMAO") %>%
  filter(SPECIES != "HAEL") #I don't know why my original code didn't work and I am too tired to fix my putrid whoopsie
```


# 1) 

## How many observations are there of each bird species in each experimental treatment? 
```{r}
raw %>%
  count(SPECIES, Rat_Removal, sort = TRUE)
```


## Make a boxplot or violinplot to show the distribution of foraging heights of each bird species in each experimental treatment.

```{r}
boxplot1 <- raw %>%
  ggplot(aes(x = SPECIES, y = foraging.ht.m, color = Rat_Removal))+
  geom_violin() +
  geom_dotplot(dotsize= 0.2, binaxis='y', stackdir='center',
                 position=position_dodge(1), binwidth = .75) +
  theme_bw()
  boxplot1
```




## What have you learned so far?

So far I've learned there's 6 bird species total observed, and that the foraging height for all birds ranges from about 2 to about 27 meters. There are also quite a lot of apapane observations, and not a lot of 'Oma'o or ... I assume HAEL is the Hawaiian Elapaio now okay I got it. It's probably not worth looking at 'oma'o or elapaio here anyway I bet. Should I remove them and just do the species you listed? Maybe? AH!- Note. I did.


# 2)

## Now make a plot showing the mean foraging height of each bird species in each treatment. include error bars displaying +/- one standard error of the mean.

```{r}
#Summary table (so I can look at it and really give myself a high five)
summary <- raw %>%
    group_by(SPECIES,Rat_Removal) %>%
  summarize(mean_height = mean(foraging.ht.m),
            n = n(), #this takes the number of observations and names it n (527)
            sd = sd(foraging.ht.m)) %>%
  mutate(se=sd/sqrt(n))


#Make a plot from the summary

meanplot2 <- summary %>%
ggplot(aes(x = SPECIES, y= mean_height, color = Rat_Removal))+
  geom_pointrange(aes(x=SPECIES, ymin=mean_height-se, ymax=mean_height+se), alpha=0.9, size=0.5,position= position_dodge(width=0.5)) +
  theme_bw()


meanplot2
```

## What is the meaning of the standard error of the mean? 
The standard error of the mean tells us how much the sample mean varies around the whole population mean. Compare to standard deviation, which tells how much the data in a sample varies around its own mean. 

## How does this plot differ from the plot in #1?
This plot makes it much clearer that, when areas are treated for rats, the average forage height for birds is lower. 


# 3) Fit a linear model 

test whether the bird species differ in foraging height, whether the rat removal treatment affects foraging height, and whether the effect of the rat removal differs between the bird species. 

```{r}
# Model 
rat.species.model <- lm(foraging.ht.m ~ Rat_Removal*SPECIES, data = raw)
#Summary 
summary(rat.species.model)
```

## Assess Residual plots
```{r}
#Residual plots 
resid_panel(rat.species.model, plots = c('resid', 'qq', 'lev', 'hist'))
```
I think... this... looks... fine. I am a little worried about how to interpret the  HOWEVER, I'm not married to that viewpoint. 




## Report F-Tests for the terms in the model
(note that if the model contains multiple predictors, these are partial residuals)
```{r}
#F stat
Anova <- Anova(rat.species.model)
Anova
```
OK. I am unsure about F-tests at this point. Not only how to report them, but I am not sure yet what they are for and now to read them despite reading things all week. BUT I WILL LEARN. Would I do, uh, F 3,498 = 0.4, p = 0.699 showing that rat removal did not affect bird foraging height differently by species? 


## Create an effects plot displaying the fitted effects
-- magnitudes of the modeled relationships-- look at the slope coefficent for linear regression.- I am confused by this hint.  

```{r}
#Effects Plot (?)
plot(ggeffect(rat.species.model, terms = c('SPECIES','Rat_Removal')))
```



## How do you interpret the results? What are the magnitudes of the effects?
 
AH. Okay. Er. No rat means bird go lower for food. The species have very slightly different foraging height ranges, but the rat removal didn't affect one species more than others. I am not sure how to assess magnitude of the effects here yet.  
 
 
 
 
# 4) Make a new model that adds kipuka area and arthopod biomass as predictors

##  Kipuka area and arthopod biomass are both continuous predictors – before you add them to the model, make some scatterplots to assess whether these predictors should be transformed when including them in the model.

When assessing predictors for transformation, it doesn’t matter if the predictors are normally
distributed (this only matters for the response variable) – *what matters is whether
the predictors are very skewed,* such that a few outlying points will have a large
influence on a fitted regression line. For skewed predictors, a log or square root
transformation will generally help.

```{r}
#scatterplot showing raw predictors
scatter <- raw %>%
  ggplot(aes(x = Area_ha, y = tot.arth.bm, color = Rat_Removal))+
  geom_point()+
  theme_bw()
  scatter
#scatterplot showing log transformed predictors
scatterlog <- raw %>%
  ggplot(aes(x = log(Area_ha), y = log(tot.arth.bm), color = Rat_Removal))+
  geom_point()+
  theme_bw()
  scatterlog
  
```
That does look better! I think? 

## Make a new model that adds kipuka area and arthopod biomass as predictors, and include interactions that test whether the treatment effect depends on these additional predictors.

So far we have examined the effect of the rat removal, but we have not considered
the role of kipuka size, or the role of food availability (which is influenced in part
by kipuka size). 
```{r}
# Model 
kipuka.model <- lm(foraging.ht.m ~ Rat_Removal * log(Area_ha) * log(tot.arth.bm), data = raw)
#Summary 
summary(kipuka.model)
```






## Report F-tests and effects plots. 
```{r}

#Residual plots 
resid_panel(kipuka.model, plots = c('resid', 'qq', 'lev', 'hist'))

#F test: 
Anova(kipuka.model)

```
 Right now, I think reporting this looks like F 1,498 = 45, p= 5x10^-11 that supports rat removal, area and biomass all influence foraging height? Do I report F tests for anythingn with a significant p value? 


```{r}
#Effects Plot 
plot(ggeffect(kipuka.model, terms = c('tot.arth.bm', 'Rat_Removal')))
plot(ggeffect(kipuka.model, terms = c('Area_ha', 'Rat_Removal')))
```

## How do the results of this model differ from the model in #3? 
Did I do this right? Is any of this right? Is this on? 

Er, I don't think I did this correctly, but it does look like the more arthropod biomass you accumulate, there is a fairly large separation between foraging areas between treated and untreated plots. Treated plots mean that birds will trend to forage lower than birds in untreated plots. I don't know if my effects plot is great since the scale seems kind of weird to me. Also, I may have done all of this very wrong. 

## How do you interpret the results at this stage?

It looks like arthropod biomass and kipuka size effect the response variable in treatment groups.  




# 5) Add DATE as a factor to the linear model from #4.
```{r}
date.model <- lm(foraging.ht.m ~ Rat_Removal * log(Area_ha) * log(tot.arth.bm) + DATE, data = raw)
#Summary 
summary(date.model)
```


## How does the inclusion of the date of sampling alter the model results? 
I am not confident enough I did the formula correctly to really make a good guess. I have a growth mindset and I will LEARN I SWEAR. 

## Why do you think that is?

Date might parse out pseudo-replication. 
Stick a fork in me! I should have signed up for baby math. 